Crime in Los Angeles: Homicide Increases in the COVID-19 Era

### 2025-04-16 Slides/Slides_2.qmd
final slides version and PDF

@@

@@

# creating labels
levels(homicide_trends$income_quartile) <- c("Low Income", "Lower-Middle Income", "Higher-Middle Income", "High Income")
# Plot homicide trends over time by income quartile
ggplot(homicide_trends, aes(x = year, y = homicide_count)) +
  geom_area(aes(fill = as.factor(income_quartile)), alpha = 0.6, position = "identity") +
  labs(
    x = "Year",
    y = "Number of Homicides",
    fill = "Income Quartile"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "plasma", direction = -1)
```

## Homicides and Median Income By Zip Code
```{r, echo=FALSE}
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(), normalizePath("crime_in_la.db"))

@@

@@

:::::: columns
:::: {.column width="35%"}
Year is significant at a 0.01 level. Each year, homicide rates increase by 0.05 on average.
::::

::: {.column width="65%"}
```{r}
con <- dbConnect(RSQLite::SQLite(), normalizePath("la_stars.db"))

query <- dbSendQuery(con, "SELECT * FROM crime_la_annually WHERE `Crm.Cd.Desc` == 'CRIMINAL HOMICIDE'")
homicides_annual <- dbFetch(query,n=-1)
dbClearResult(query)
dbDisconnect(con)

ggplot(homicides_annual, aes(x = year, y = variable)) +
  geom_col(fill = "#19297C") +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_text(), title = element_text(size = 12)) +
  labs(title="Number of Homicides in LA, Annually",
       x = "Year",
       y = "Homicide Count")
```
:::
::::::

@@

```{r}
#| context: setup
#| include: false
# load functions
source("dependencies.R")
library(dplyr)

# load data
con <- dbConnect(RSQLite::SQLite(), normalizePath("la_stars.db"))

@@

map_datasets <- list("crime" = crime_la_annually,
                     "covid" = covid_la_annually,
                     "evictions" = evictions_county_la,
                     "budget" = lapd_budget_annually,
                     "unemploy" = unemployment_la)

@@

```{r}
#| panel: sidebar
  # Dropdown to select dataset
  selectInput("dataset", "Choose Dataset", 
              choices = c("Crime" = "crime", "COVID" = "covid", "Evictions" = "evictions", "LAPD Budget" = 'budget', "Unemployment" = 'unemploy'))

sliderInput("year_range", "Select Time Range:",
              min = 2010, max = 2025, value = c(2015, 2020), sep = "")  # Placeholder, updated dynamically
uiOutput("crime_type_ui")
```

@@

selectedKillerData <- reactive({
  if (input$dataset == 'crime') {
    map_datasets[[input$dataset]] %>% filter(year %in% seq(input$year_range[1], input$year_range[2]),
                                             Crm.Cd.Desc == input$crime_type)
  }

  else {
    map_datasets[[input$dataset]] %>% filter(year %in% seq(input$year_range[1], input$year_range[2]))
  }

  })

observeEvent(input$dataset, {
    # Get year vector based on dataset selected
    year_vals <- switch(input$dataset,
                        "crime" = crime_la_annually$year,
                        "covid" = covid_la_annually$year,
                        "evictions" = evictions_county_la$year,
                        "budget" = lapd_budget_annually$year,
                        "unemploy" = unemployment_la$year)

    # Update slider with dynamic range
    updateSliderInput(session, "year_range",
                      min = min(year_vals),
                      max = max(year_vals),
                      value = c(min(year_vals), max(year_vals)))
  })

output$crime_type_ui <- renderUI({
    if (input$dataset == "crime") {
      selectInput("crime_type", "Select Crime Type",
                  choices = unique(crime_la_annually$Crm.Cd.Desc),
                  selected = unique(crime_la_annually$Crm.Cd.Desc)[1])
    }
  })

output$stars <- renderPlot({
      df <- selectedKillerData()
      print(df)
      rows <- nrow(df)

      col_row <- if (rows <= 13) {
        c(rows, 1)
      } else if (rows == 14) {
        c(7, 2)
      } else if (rows == 15) {
        c(5, 3)
      } else if (rows == 16) {
        c(4, 4)
      } else {
        c(ceiling(sqrt(rows)), ceiling(rows / ceiling(sqrt(rows))))
      }

      stars <- generate_star_centers(col_row[1], col_row[2])
      labels <- levels(standardize_scale(df$variable))
      print(labels)
      df_sorted <- df[order(df$year), ]

@@

        grid.text(label = as.character(row$year),
    x = unit(row$x, "npc"),
    y = unit(row$y - 0.05, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 10))

        grid.rect(x = 0.5, y = 0, width = 1, height = 0.125, just = "bottom", gp = gpar(col = "white", fill = "gray15"))

        grid.text(label = paste(stringr::str_to_title(input$dataset), "Level"),
    x = unit(.1, "npc"),
    y = unit(0.0625, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 14))
      }

      grid.draw(create_star(scale_factor = 1, center = c(.2, .0625)))

@@

      grid.text(label = as.character(labels[2]), x = unit(.275, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[3]), x = unit(.475, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[4]), x = unit(.725, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[5]), x = unit(.925, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
    })
```

>>>>

## Weekly Crime Counts in Los Angeles from 2010 to present

query <- dbSendQuery(
  con,
  "
  SELECT
    YearWeek,
    MIN(Date) AS start_of_week,
    MAX(Date) AS end_of_week,
    COUNT(*)  AS Crimes
  FROM crime_la
  GROUP BY YearWeek
  ORDER BY YearWeek
  "
)
weekly_counts <- dbFetch(query, n = -1)
dbClearResult(query)
dbDisconnect(con)

weekly_counts$start_of_week <- as.Date(weekly_counts$start_of_week)
weekly_counts$end_of_week   <- as.Date(weekly_counts$end_of_week)

weekly_counts$days_in_week <-
  as.numeric(difftime(weekly_counts$end_of_week,
                      weekly_counts$start_of_week,
                      units = "days")) + 1

weekly_counts <- subset(weekly_counts, days_in_week == 7)

weekly_counts$CrimesScaled <-
  weekly_counts$Crimes * (7 / weekly_counts$days_in_week)

weekly_counts$Index <- seq_len(nrow(weekly_counts))

march_2020_index <- which(weekly_counts$YearWeek == "2020-09")

plot(
  weekly_counts$Index,
  weekly_counts$CrimesScaled,
  type  = "l",
  col   = "blue",
  lwd   = 2,
  xlab  = "Weekly Index",
  ylab  = "Number of Crimes (Scaled to 7 Days)"
)

abline(
  h   = pretty(weekly_counts$CrimesScaled),
  v   = pretty(weekly_counts$Index),
  col = "gray",
  lty = "dotted"
)

abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)

text(
  x     = weekly_counts$Index[march_2020_index] - 80,
  y     = 10,
  labels = "March 2020",
  pos   = 3,
  col   = "red"
)

```

@@

```{r, results='asis'}
library(plotly)
con <- dbConnect(RSQLite::SQLite(), normalizePath("crime_in_la.db"))

@@

# creating labels
levels(homicide_trends$income_quartile) <- c("Low Income", "Lower-Middle Income", "Higher-Middle Income", "High Income")

homicide_trends <- homicide_trends %>%
  mutate(income_quartile = as.factor(income_quartile)) %>%
  rename("Year" = "year",
         "Homicide Count" = "homicide_count",
         "Income Quartile" = "income_quartile"
         )
# Plot homicide trends over time by income quartile
p <- ggplot(homicide_trends, aes(x = Year, y = `Homicide Count`)) +
  geom_area(aes(fill = `Income Quartile`), alpha = 0.6, position = "identity") +
  labs(
    x = "Year",
    y = "Number of Homicides",
    fill = "Income Quartile"
  ) +
  theme_minimal() +
  scale_fill_viridis_d(option = "plasma", direction = -1)

ggplotly(p)
```

## Homicides and Median Income By Zip Code

```{r, echo=FALSE}
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(), normalizePath("crime_in_la.db"))

@@

## Relationship Between Homicides and Socioeconomic Variables
```{r}
#| context: setup
#| include: false
library(RSQLite)
library(ggplot2)
library(dplyr)

corr <- read.csv("corr.csv")

colnames(corr) <- c("X", "Year", "Month", "Unemployment", "Crimes", "COVID", "Evictions", "LAPD Budget")
```

```{r}
#| panel: sidebar
radioButtons("opts", "View homicides compared to:", 
            choices = c("COVID", "Evictions", "LAPD Budget", "Unemployment"))
```

```{r}
#| panel: fill
plotOutput("scatter")
```

```{r}
#| context: server
output$scatter <- renderPlot({
  # Guessing 'value' is the comparison variable — adjust as needed
  df <- corr[c("Crimes", input$opts, 'Year')]
  colnames(df) <- c("Crimes", "var", "Year")

  ggplot(df, aes(x = var, y = Crimes)) +
    geom_point(aes(colour = as.factor(Year))) +
    scale_color_viridis_d() +
  ggthemes::theme_fivethirtyeight() +
  theme(axis.title = element_text()) + 
    labs(x = input$opts, y = "Crime Count", title = paste("Monthly Crime vs", input$opts),
         colour = "Year") +
    geom_smooth(method = "lm")
}) 
```

@@

Year is significant at a 0.01 level, but it has caused a very small increase, almost unnoticeable on the graph. Each year, homicide rates increase by 0.05 on average.

@@

```{r}
#| context: setup
#| include: false
# load functions
source("dependencies.R")
library(dplyr)
spatial <- read.csv("homicides.csv")
# load data
con <- dbConnect(RSQLite::SQLite(), normalizePath("la_stars.db"))

@@

map_datasets <- list("Crime Count" = crime_la_annually,
                     "Covid Cases" = covid_la_annually,
                     "Evictions" = evictions_county_la,
                     "LAPD Budget" = lapd_budget_annually,
                     "Unemployment Rate" = unemployment_la)

@@

```{r}
#| panel: sidebar

radioButtons("type", "View:", choices = c("Homicides Spatially", "A Variable Over Time"), selected = "Homicides Spatially")
# Dropdown to select dataset
uiOutput("dataset")

uiOutput("year_selector")

uiOutput("crime_type_ui")
```

@@

selectedKillerData <- reactive({
  req(input$dataset, input$year_range)

  if (input$dataset == 'Crime Count') {
    req(input$crime_type)
    df <- map_datasets[[input$dataset]] %>%
      filter(year %in% seq(input$year_range[1], input$year_range[2]),
             Crm.Cd.Desc == input$crime_type)
  } else {
    df <- map_datasets[[input$dataset]] %>%
      filter(year %in% seq(input$year_range[1], input$year_range[2]))
  }
  return(df)
})

output$dataset <- renderUI({
  req(input$type)

  if (input$type == "A Variable Over Time") {
  selectInput("dataset", "Choose Dataset", 
            choices = c("Crime Count", 
                        "Covid Cases", 
                        "Evictions", 
                        "LAPD Budget", 
                        "Unemployment Rate")) 
}
})

output$year_selector <- renderUI({
  req(input$type)

  if (input$type == "Homicides Spatially") {
    # Show this when Homicides Spatially is selected
    selectInput("year_single", "Select Year:", choices = 2019:2023, selected = 2023)
  } else {
    req(input$dataset)  # Now only required when viewing variables over time

    # Get relevant years
    year_vals <- switch(input$dataset,
                        "Crime Count" = crime_la_annually$year,
                        "Covid Cases" = covid_la_annually$year,
                        "Evictions" = evictions_county_la$year,
                        "LAPD Budget" = lapd_budget_annually$year,
                        "Unemployment Rate" = unemployment_la$year)

    year_vals <- sort(unique(year_vals))

    sliderInput("year_range", "Select Time Range:",
                min = min(year_vals), max = max(year_vals),
                value = c(min(year_vals), max(year_vals)), step = 1,
                sep = "")
  }
})

output$crime_type_ui <- renderUI({
  req(input$type)
  req(input$dataset)

  if (input$type == "A Variable Over Time" && input$dataset == "Crime Count") {
    selectInput("crime_type", "Select Crime Type",
                choices = unique(crime_la_annually$Crm.Cd.Desc),
                selected = unique(crime_la_annually$Crm.Cd.Desc)[1])
  }
})

output$stars <- renderPlot({
  req(input$type)

  if (input$type == "Homicides Spatially") {
    req(input$year_single)
    print(as.integer(input$year_single))
    print(unique(spatial$year))
    df <- spatial %>% filter(year == as.integer(input$year_single))
    df$LAT <- round(df$LAT_rounded, 1)
    df$LON <- round(df$LON_rounded, 1)

    df <- df %>% group_by(LAT, LON) %>% summarize(avgHomicides = mean(HomicideCount)) %>% ungroup()

    df$LATstd <- .175 + ((df$LAT - min(df$LAT))/(max(df$LAT) - min(df$LAT))) * .75
    df$LONstd <- .175 + ((df$LON - min(df$LON))/(max(df$LON) - min(df$LON))) * .75
   labels <- levels(standardize_scale(df$avgHomicides))
   df$scale <- as.numeric(standardize_scale(df$avgHomicides))

   grid.newpage()
   grid.rect(gp = gpar(fill = "gray15", col = NA))  # No border, black fill
   for (star in 1:nrow(df)) {
        row <- as.list(df[star,])
        grid.draw(create_star(scale_factor = row$scale, center = c(row$LONstd, row$LATstd))) 
        grid.text(label = paste0(row$LAT, ", ", row$LON),
    x = unit(row$LONstd, "npc"),
    y = unit(row$LATstd + 0.05, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 10))
        }
   grid.rect(x = 0.5, y = 0, width = 1, height = 0.125, just = "bottom", gp = gpar(col = "white", fill = "gray15"))

        grid.text(label = "Average Homicides",
    x = unit(.125, "npc"),
    y = unit(0.0625, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 14))

      grid.draw(create_star(scale_factor = 1, center = c(.25, .0625)))

      grid.text(label = as.character(labels[1]), x = unit(.275, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[2]), x = unit(.475, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[3]), x = unit(.725, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[4]), x = unit(.925, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11)) 
} else {
      df <- selectedKillerData()
      print(head(df))
      rows <- nrow(df)
      print(rows)
      col_row <- if (rows <= 13) {
        c(rows, 1)
      } else if (rows == 14) {
        c(7, 2)
      } else if (rows == 15) {
        c(5, 3)
      } else if (rows == 16) {
        c(4, 4)
      } else {
        c(ceiling(sqrt(rows)), ceiling(rows / ceiling(sqrt(rows))))
      }

      stars <- generate_star_centers(col_row[1], col_row[2])
      print(nrow(stars))
      labels <- levels(standardize_scale(df$variable))
      print(labels)
      df_sorted <- df[order(df$year), ]

@@

        grid.text(label = as.character(row$year),
    x = unit(row$x, "npc"),
    y = unit(row$y - 0.05, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 10))
      }
        grid.rect(x = 0.5, y = 0, width = 1, height = 0.125, just = "bottom", gp = gpar(col = "white", fill = "gray15"))

        grid.text(label = paste(stringr::str_to_title(input$dataset), "Level"),
    x = unit(.1, "npc"),
    y = unit(0.0625, "npc"),  # offset below star; adjust as needed
    gp = gpar(col = "white", fontsize = 14))

      grid.draw(create_star(scale_factor = 1, center = c(.225, .0625)))

@@

      grid.text(label = as.character(labels[2]), x = unit(.275, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[3]), x = unit(.475, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[4]), x = unit(.725, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11))
      grid.text(label = as.character(labels[5]), x = unit(.925, "npc"),
                y = unit(0.0625, "npc"), gp = gpar(col = "white", fontsize = 11)) }
    })
```
